HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-15 scnn.tex
fix usage of i.e.

The first approach, leveraging the continuous spherical convolution, has been
proposed by~\cite{cohen2018spherical}. The goal of
these contributions it to use a rotational equivariant convolution on the
sphere, i.e. a rotation of the input implies the same rotation of the output.
The resulting convolution is performed by a spherical Fourier transform (i.e.
a projection on the spherical harmonics), a  multiplication in the spectral
domain and an inverse spherical Fourier transform. Hence the main
computational cost of a convolution comes from the two Fourier transforms.
Fortunately, the Fourier transform can be accelerated for special samplings
set (theoretically including HEALPix, \cite{mohlenkamp1999fast,rokhlin2006fast}).
The main advantage of this
approach is that it provides a mathematically well defined rotational
equivariant network. Nevertheless, even with these Fourier transform
acceleration, the convolutions remain expensive in comparison to a traditional
2-dimensional convolution, limiting the practical use of this approach.
\todo{to have numbers to back up our claim: have the theoretical complexities and measure the time needed for a convolution on graph and the healpix Fourier transform to get the constants
\nati{@mdeff: for the complexity we probably have $\mathcal{O}(n)$ instead of $\mathcal{O}(n^{3/2})$ (let assume that we have $\approx n^{1/2}$ points allong both axis, computation of the SO3 FFT is composed of 1) $n^{1/2}$ FFTs for a total cost of $n \log (n)$ and 2) $n^{1/2}$ matrix multiplications of size $n^{1/2}$ for a total cost of $n^{3/2}$) For the number, it is a bit more complicated… I guess it is for the next paper.}
}
While a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the HEALPix one. Hence, a practical comparison is
not possible. We also note that this approach cannot be trivially accelerated
if only one part of the sphere is used. This is impractical in the field of
cosmology as the data often observed only on a subpart of the sphere.
One could still pad the unused part of the sphere with 0. However, this would significantly increase the computational cost of the method.
\nati{@mdeff: do you think we can say the method would be too slow for our dataset even without trying it? I am not so sure...}
\todo{slow: full sky maps are composed of millions of points. partial maps are hard to process}

A second direction has been followed in~\citet{boomsma2017spherical}. The idea is
to use the traditional 2-dimensional convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the equirectangular grid and the ``cubed
sphere'' grid defined by~\citet{ronchi1996cubed}. \nati{Do we need a plot of the equirectangular and cubed sphere sampling?} Because this
approach is based on the traditional 2-dimensional convolution, it is very likely to be the
most computationally efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are not including HEALPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e. the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere.
\todo{border effects? \nati{@mdeff what are you thinking about here?}}
% \nati{Michael: Under some very
% specific hypothesis, this second approach is a particular case of our method,
% i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
% on a grid == 2d convolution. Do you think we should mention that? I think we
% should not.}
% I think we should!

@@

\subsection{Convolutions on graphs}
\todo{Maybe we just describe convolutions on sphere and graph, i.e: do we really need all these three subsections? Maybe simply merge them?}

@@

We build a robust classifier, i.e. an automatic way to
discriminate between the two classes in the presence of noise.
We check the classification performance as a function of the area used, i.e "the order" $o$ and also as a function of the relative level of the noise added to the map.

@@

We present in \figref{results} the results for 5 different noise
levels and $3$ different sample areas. The spherical CNN has a significant lead for all levels of noise and sample areas.
In particular, the difference seems to increase as the problem becomes more complex, i.e as the order is increased and there is less information in each sample.
This is likely due to the fact that the SCNN can learn optimal features for the problem and do not rely on our humanly limited feature design.

We observe that the PSD features, i.e.
one of the most used techniques for model discrimination in cosmology, is unable
to perform the desired task. The main reason is that the two different models
have very similar PSDs (see \figref{psd_sigma3}). As a result, one
needs to use other statistics to solve the problem. In this case, the
histogram features that have been built for the problem allow for a significant improvement over the PSD ones.
Unfortunately, while the histogram features contain information about the
distribution of pixels, they do not include any spatial information. In
contrast, both the PSD features and the SCNN can exploit the spatial
information.
Note that without this graph structure, a fully connected neural network approach would have to be used. Problematically, fully connected neural networks do not scale well with the number of pixels.

@@

In this contribution, we presented HealpixNet, a spherical CNN for the HEALPix sampling. The gist of our construction resides in using a graph to model the sphere.
This allows us to leverage the advantages of the graph convolution, i.e it is both efficient (with a complexity of $\mathcal{O}(n)$) and flexible (allow to work only on a subpart of the sphere) while it still captures well the spherical properties of the domain.
We also show on a range of classification problems that the neural network systematically significantly outperform classical methods.

@@

In \figref{gaussian_filters_visualization}, we present three different
visualization for a HEALPix graph convolution kernel. The kernel used is defined in the graph spectral domain as $K_t(x)=e^{-\tau t x}$. This is the domain used for our first visualization. In this case, we simply evaluate $K_t$
at the graph eigenvalues $\text{diag}(\bLambda)$. The lower eigenvalues $\lambda$ correspond to the lower frequencies, i.e. the spectral mode with less variation.
As a second visualization, we can observe the kernel directly on the sphere. To do so, we perform the convolution of a
Kroneker on the sphere plot the result with gnomonic projection. Eventually, as we are
working on the sphere with theoretically spherical filters (up to the irregularities of the sampling), we could only plot the section of the convoluted Kroneker. To do so, we convolve a Kroneker on the equator and use the equator for our section as shown in~\figref{index_section}. Note that, because of the small irregularities in the HEALPix
sampling, the second and the third methods are subject to small variations depending on the
chosen Kroneker.

>>>>

The first approach, leveraging the continuous spherical convolution, has been
proposed by~\cite{cohen2018spherical}. The goal of
these contributions it to use a rotational equivariant convolution on the
sphere, i.e., a rotation of the input implies the same rotation of the output.
The resulting convolution is performed by a spherical Fourier transform (i.e.,
a projection on the spherical harmonics), a  multiplication in the spectral
domain and an inverse spherical Fourier transform. Hence the main
computational cost of a convolution comes from the two Fourier transforms.
Fortunately, the Fourier transform can be accelerated for special samplings
set (theoretically including HEALPix, \cite{mohlenkamp1999fast,rokhlin2006fast}).
The main advantage of this
approach is that it provides a mathematically well defined rotational
equivariant network. Nevertheless, even with these Fourier transform
acceleration, the convolutions remain expensive in comparison to a traditional
2-dimensional convolution, limiting the practical use of this approach.
\todo{to have numbers to back up our claim: have the theoretical complexities and measure the time needed for a convolution on graph and the healpix Fourier transform to get the constants
\nati{@mdeff: for the complexity we probably have $\mathcal{O}(n)$ instead of $\mathcal{O}(n^{3/2})$ (let assume that we have $\approx n^{1/2}$ points allong both axis, computation of the SO3 FFT is composed of 1) $n^{1/2}$ FFTs for a total cost of $n \log (n)$ and 2) $n^{1/2}$ matrix multiplications of size $n^{1/2}$ for a total cost of $n^{3/2}$) For the number, it is a bit more complicated… I guess it is for the next paper.}
}
While a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the HEALPix one. Hence, a practical comparison is
not possible. We also note that this approach cannot be trivially accelerated
if only one part of the sphere is used. This is impractical in the field of
cosmology as the data often observed only on a subpart of the sphere.
One could still pad the unused part of the sphere with 0. However, this would significantly increase the computational cost of the method.
\nati{@mdeff: do you think we can say the method would be too slow for our dataset even without trying it? I am not so sure...}
\todo{slow: full sky maps are composed of millions of points. partial maps are hard to process}

A second direction has been followed in~\citet{boomsma2017spherical}. The idea is
to use the traditional 2-dimensional convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the equirectangular grid and the ``cubed
sphere'' grid defined by~\citet{ronchi1996cubed}. \nati{Do we need a plot of the equirectangular and cubed sphere sampling?} Because this
approach is based on the traditional 2-dimensional convolution, it is very likely to be the
most computationally efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are not including HEALPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e., the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere.
\todo{border effects? \nati{@mdeff what are you thinking about here?}}
% \nati{Michael: Under some very
% specific hypothesis, this second approach is a particular case of our method,
% i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
% on a grid == 2d convolution. Do you think we should mention that? I think we
% should not.}
% I think we should!

@@

\subsection{Convolutions on graphs}
\todo{Maybe we just describe convolutions on sphere and graph, i.e.: do we really need all these three subsections? Maybe simply merge them?}

@@

We build a robust classifier, i.e., an automatic way to
discriminate between the two classes in the presence of noise.
We check the classification performance as a function of the area used, i.e., "the order" $o$ and also as a function of the relative level of the noise added to the map.

@@

We present in \figref{results} the results for 5 different noise
levels and $3$ different sample areas. The spherical CNN has a significant lead for all levels of noise and sample areas.
In particular, the difference seems to increase as the problem becomes more complex, i.e., as the order is increased and there is less information in each sample.
This is likely due to the fact that the SCNN can learn optimal features for the problem and do not rely on our humanly limited feature design.

We observe that the PSD features, i.e.,
one of the most used techniques for model discrimination in cosmology, is unable
to perform the desired task. The main reason is that the two different models
have very similar PSDs (see \figref{psd_sigma3}). As a result, one
needs to use other statistics to solve the problem. In this case, the
histogram features that have been built for the problem allow for a significant improvement over the PSD ones.
Unfortunately, while the histogram features contain information about the
distribution of pixels, they do not include any spatial information. In
contrast, both the PSD features and the SCNN can exploit the spatial
information.
Note that without this graph structure, a fully connected neural network approach would have to be used. Problematically, fully connected neural networks do not scale well with the number of pixels.

@@

In this contribution, we presented HealpixNet, a spherical CNN for the HEALPix sampling. The gist of our construction resides in using a graph to model the sphere.
This allows us to leverage the advantages of the graph convolution, i.e., it is both efficient (with a complexity of $\mathcal{O}(N_{pix})$) and flexible (allow to work only on a part of the sphere) while it still captures well the spherical properties of the domain.
We also show on a range of classification problems that the neural network systematically significantly outperform classical methods.

@@

In \figref{gaussian_filters_visualization}, we present three different
visualization for a HEALPix graph convolution kernel. The kernel used is defined in the graph spectral domain as $K_t(x)=e^{-\tau t x}$. This is the domain used for our first visualization. In this case, we simply evaluate $K_t$
at the graph eigenvalues $\text{diag}(\bLambda)$. The lower eigenvalues $\lambda$ correspond to the lower frequencies, i.e., the spectral mode with less variation.
As a second visualization, we can observe the kernel directly on the sphere. To do so, we perform the convolution of a
Kroneker on the sphere plot the result with gnomonic projection. Eventually, as we are
working on the sphere with theoretically spherical filters (up to the irregularities of the sampling), we could only plot the section of the convoluted Kroneker. To do so, we convolve a Kroneker on the equator and use the equator for our section as shown in~\figref{index_section}. Note that, because of the small irregularities in the HEALPix
sampling, the second and the third methods are subject to small variations depending on the
chosen Kroneker.
